---
title: "R as a GIS - Basics"
author: "Paul Chapron & Hadrien Commenges"
date: "April 2019"
output: 
  html_document: 
    highlight: kate
    theme: simplex
    toc: yes
    toc_float: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE)
```


## Outline

- Loading a variety of types and formats of spatialized data
- Performing basic GIS operations
- Visualizing the results with plots and maps

## Data description

The dataset covers the Paris region.

- Fine-grain administrative boundaries (polygons, ESRI shp)
- Elevation data (raster, ASC)
- Population grid with income data (polygons, Rds)
- GPS tracks for my city walk (points, geoJSON)

## Packages and data

```{r}
library(sf)
library(raster)
library(leaflet)
library(tidyverse)
```

Load the Paris neighborhoods (Iris perimeters) and **dissolve to get one unified polygon**.

```{r}
parisNei <- st_read("DATA/Demo/parisnei.shp", crs = 2154, stringsAsFactors = FALSE)
plot(parisNei$geometry)

# option 1 (geometry union)
parisUnion <- parisNei %>% st_union()
plot(parisUnion)

# option 2 (summarizing attributes and dissolving geometries)
parisBound <- parisNei %>% 
  group_by(DEP) %>% 
  summarise(NBUNITS = n())
plot(parisBound)
```


Load the raster of elevation and **declare the coordinate reference system** (CRS).

```{r cars}
rastElev <- raster(x = "DATA/Demo/BDALTI_75.asc")
projection(rastElev) <- CRS("+init=epsg:2154")
```

Load the population grid and **convert to simple feature** format (`sf`).

```{r}
incomeGrid <- readRDS("DATA/Demo/carroyage.Rds") %>% 
  st_as_sf() %>% 
  st_set_crs(value = 2154)
st_crs(incomeGrid)
```

Load the GPS data from my city walk and **transform the coordinates reference system**.

```{r}
myRoute <-   st_read("DATA/Demo/myroute.geojson")
st_crs(myRoute)
myRouteProj <- st_transform(myRoute, crs = 2154)
```

```{r}
plot(parisNei$geometry, col = "grey", border = "white")
plot(myRouteProj$geometry, pch = 4, cex = 0.1, col = "firebrick", add = TRUE)
```


## Objects to produce

- **Object:** interactive map of the city walk with the Paris neighborhoods
- **Operations to perform:**
  1. Create a line from the collection o points [`st_linestring()`, `st_sf()`]
  2. Re-project the Paris neighborhoods [`st_transform()`]
  2. Map the neighborhoods and the line on OSM tiles [`leaflet()`, `addTiles()`, `addPolygons()`, `addPolylines()`]

- **Object:** static map of the city walk with the income grid (median income as choropleth)
- **Operations to perform:**
  1. Create a line from the collection o points [`st_linestring()`, `st_sf()`]
  2. Cut or mask the population grid with a relevant bounding polygon or bounding box [`st_bbox()`, `st_intersects()` or `st_contains()` or `st_crop()`]
  3. Create a choropleth layer with the median income for each cell of the grid with the `cartography` package [`choroLayer()`]
  4. Add the spatial line representing the city walk [`plot()`]

- **Object:** Elevation profile for my city walk
- **Operations to perform:**
  1. extract elevation values from the raster for each GPS point [`raster::extract()`, !warning! `tidyr::extract()`]
  2. compute pairwise distances between successive points [`st_distance()`]
  3. plot the result: elevation (Y) vs. distance (X)

- **Object:** Income profile for my city walk
- **Operations to perform:**
  1. extract income values from the grid for each GPS point [`st_intersects()`]
  2. compute pairwise distances between successive points [`st_distance()`]
  3. plot the result: elevation (Y) vs. distance (X) [`plot()` or `ggplot()`]



